One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-09 paper/discussion.tex
intro draft completed

\section{Discussion}
Having confirmed the $m=1$ spiral instability through numerical
simulations, we now use results from the simulations combined with
linear theory to construct an explanation for their growth. For
simplicitly, we return to 2D discs.

>>>>

\section{Discussion}\label{discussions}
Having confirmed the $m=1$ spiral instability through numerical
simulations, we now use results from the simulations combined with
linear theory to construct an explanation for their growth. For
simplicitly, we return to 2D discs.

### 2014-11-09 paper/paper.tex
intro draft completed

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJL}
\newcommand{\pasj}{PASJ}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\bu}{\bm{u}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\teta}{\tilde{\eta}}
\newcommand{\tcool}{t_\mathrm{cool}}
\newcommand{\jlin}{j_\mathrm{lin}}
\newcommand{\jlintot}{J_\mathrm{lin}}
\newcommand{\tbg}{T_\mathrm{BG}}
\newcommand{\varA}[1]{{\operatorname{#1}}}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\real}{Re}
\DeclareMathOperator{\imag}{Im}

@@

\section{Introduction}\label{intro}
% Protoplanetary discs are expected to contain complex structures
% \citep{armitage11}, which directly impact planet formation theories.  
One of the exciting developments in the study of circumstellar 
discs is the direct observation of large-scale, non-axisymmmetric
structures within them. These
include lopsided dust distributions 
\citep{marel13,fukagawa13,casassus13,isella13,perez14,follette14} and
spiral arms 
\citep{muto12,grady13,christiaens14,avenhaus14,boccaletti14}.

% planet is popular answer for spiral structure
An attractive explanation for asymmetries in circumstellar discs is to
invoke disc-planet interaction. In particular, spiral structures
naturally arise from the gravitational interaction between a
planet and the gaseous protoplanetary disc it is embedded in
\citep[see, e.g.][for a recent review]{baruteau13b}. Thus, the
presence of spiral arms in circumstellar discs could be signposts of 
planet formation.

% but also natural in GI.
Spiral arms are also naturally associated with global gravitational 
instability (GI) of differentially rotating discs
\citep{goldreich65,laughlin96b,laughlin98,lodato05,forgan11}. Large-scale
spiral arms may provide significant angular momentum transport to enable
accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
spiral structures are potentially observable with the Atacama Large 
Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}.

@@

% original motivation for this work (DZGI) 
Indeed, the original goal for this project was to clarify
what type of GI is relevant to astrophysical discs in which only part 
of it is self-gravitating, such as a protoplanetary disc containing a 
massive dead zone. However, our seemingly simple initial disc models
consistently produced one-armed spiral structures, despite not 
satisfying the usual requirements to generate them
(see below). We therefore dedicate this study to verify their origin.

%brief history of m=1 
%stellar wobble from indiana group 
Single-arm, or eccentric modes in

%this work  mechanism is not strictly sg

%this paper is organised as follows

>>>>

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\aj}{AJ}
\newcommand{\aap}{A\&A}
\newcommand{\apss}{ApSS}
\newcommand{\apjl}{ApJL}
\newcommand{\pasj}{PASJ}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\bu}{\bm{u}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\teta}{\tilde{\eta}}
\newcommand{\tcool}{t_\mathrm{cool}}
\newcommand{\jlin}{j_\mathrm{lin}}
\newcommand{\jlintot}{J_\mathrm{lin}}
\newcommand{\tbg}{T_\mathrm{BG}}
\newcommand{\varA}[1]{{\operatorname{#1}}}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\real}{Re}
\DeclareMathOperator{\imag}{Im}

@@

\section{Introduction}\label{intro}
% Protoplanetary discs are expected to contain complex structures
% \citep{armitage11}, which directly impact planet formation theories.  
An exciting development in the study of circumstellar 
discs is the direct observation of large-scale, non-axisymmmetric
structures within them. These
include lopsided dust distributions 
\citep{marel13,fukagawa13,casassus13,isella13,perez14,follette14} and
spiral arms 
\citep{muto12,grady13,christiaens14,avenhaus14,boccaletti14}.

% planet is popular answer for spiral structure
The attractive explanation for asymmetries in circumstellar discs is to
invoke disc-planet interaction. In particular, spiral structures
naturally arise from the gravitational interaction between a
planet and the gaseous protoplanetary disc it is embedded in
\citep[see, e.g.][for a recent review]{baruteau13b}. Thus, the
presence of spiral arms in circumstellar discs could be signposts of 
planet formation.

% but also natural in GI.
However, spiral arms are also naturally associated with global gravitational 
instability (GI) of differentially rotating discs
\citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale
spiral arms may provide significant angular momentum transport to enable
accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
spiral structures due to GI are potentially observable with the Atacama Large 
Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}.

@@

% original motivation for this work (DZGI) 
Given the wide range of GIs that may occur in  gaseous 
astrophysical discs \citep{papaloizou89,christo92,christo93,hadley11,hadley14}, it would of
theoretical interest to clarify what type of GI is relevant to 
partially self-gravitating discs such as a protoplanetary disc containing a 
massive dead zone. This was indeed the original goal of this project. 
However, our seemingly simple initial disc models
consistently produced \emph{one-armed} spiral structures, despite not 
satisfying the usual requirements to generate them 
(see below). We therefore dedicate this study to verify their origin.

%brief history of m=1 
%stellar wobble from indiana group 
Single-arm spirals, or eccentric modes, corresponding to perturbations 
with azimuthal wavenumber $m=1$, have received interest in
the context of protoplanetary discs because of their global nature
\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. Their development is 
often associated with the unique property of $m=1$ perturbations to
move the central star \citep{shu90}, which may have observable
consequeces \citep{michael10}. 
%these are self-gravitating instabilities

The aforementioned analytical studies have mostly adopted barotropic, smooth disc models. 
In this work, we identify a mechanism to generate one-arm
spiral patterns that is related to a non-barotropic equation of
state and disc structure. Specifically, we find that if the disc temperature or
sound-speed is forced to be a prescribed function of position, then 
one-arm spiral modes grows inside a cavity that results from the
adopted disc structure.

Our study is based on direct hydrodynamic simulations using three
different grid-based codes, both in two-dimensions (2D) and
three-dimensions (3D). We also apply results from previous linear
theory to explain their growth. We find that, when the disc
temperature is prescribed, the usual statement for the conservation of
angular momentum for linear perturbations acquires a source term
proportional to the temperature gradient which may act to destabilise
$m=1$ perturbations. Although we work with self-gravitating discs, the
source term itself does not depend explicitly on disc self-gravity. Thus, the
destabilisation effect we describe may also operate in
non-self-gravitating discs.

This paper is organised as follows. In \S\ref{model} we list the
governing equations and describe the disc equilibria under
consideration. \S\ref{methods} describes the hydrodynamic codes
employed in this work and several diagnostic measures. We briefly
review basic results from linear theory in \S\ref{wkb} to aid our
interpretation of numerical simulations.  Our main results are presented in 
\S\ref{results2d} for 2D simulations and \S\ref{results3d} for 3D
simulations, and we further analyse them in \S\ref{discussions}. 
We summarise in \S\ref{summary}, including a discussion on
spectulations for future work.
